Elo World,

### 2019-03-11 chess/papers/weak.tex
Better progress updates in Tournament. About to do incremental rewrite
Tournament now accumulates results to database. Separated elo/table code out into its own program.
Add fatalist, survivalist, equalizer. improve Player interface and ability to select games for tournament
Did some writing on weak.tex

% Dramatic failure of the "suicide king" strategy against
% CCCP.
1. d4 g5 2. Bxg5 Nc6 3. Bxe7 Kxe7 4. d5 Kd6
5. dxc6+ Kc5 6. Qd6+ Kc4 7. e4#

>>>>

% [computermodern], 
\documentclass[numbers]{sigplanconf}

\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}

% \usepackage{chessfs}
\usepackage{adjustbox}

% Define black versions of pieces for inline use. Gross, but it works.
\newcommand{\Pawn}[1][1.3ex]{%
\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
}%
\newcommand{\Rook}[1][1.58ex]{%
\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
}%
\newcommand{\Knight}[1][1.85ex]{%
\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
}%
\newcommand{\Bishop}[1][1.79ex]{%
\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
}%
\newcommand{\Queen}[1][2.05ex]{%
\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
}%
\newcommand{\King}[1][1.95ex]{%
\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
}%

\interfootnotelinepenalty=0

% lets me explicitly set a. or 1. etc. as enum label
\usepackage{enumitem}

\pagestyle{empty}

\usepackage{ulem}
% go back to italics for emphasis, though
\normalem

\usepackage{natbib}

\setlength{\footnotesep}{2em}

% \newcommand\comment[1]{}
\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

\begin{document}

\conferenceinfo{SIGBOVIK~2019}{April 1, 2019, Pittsburgh, Pennsylvania, USA}
\copyrightyear{2019}
\title{Elo World,}
\subtitle{a framework for benchmarking weak chess engines}
\authorinfo{Dr. Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
\toappear{
Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
Appears in SIGBOVIK 19119 with the
%
title inflation
%
of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1600 rating points}

\ccsdesc[500]{Evaluation methodologies~Tournaments}
\ccsdesc[500]{Chess~Being bad at it}
% \ccsdesc[100]{Software and its engineering~Control structures}
% \ccsdesc[300]{Theory of computation~Control primitives}

\terms
pawn, horse, bishop, castle, queen, king
\keywords
chess, bad

\setchessboard{showmover=false}

\newcommand\checkmate{\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}

\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
\newcommand\at{\ensuremath{\scriptstyle @}}

\date{1 April 2019}

\maketitle \thispagestyle{empty}

\section{Introduction}

Fiddly bits aside, it is a solved problem to maintain a numeric rating
of players for some game (for example chess, but also sports,
e-sports, probably also z-sports if that's a thing). Though it has
some competition (suggesting the need for a meta-rating system to
compare them), the Elo Rating System~\cite{elo1978rating} is a simple
and effective way to do it. This paper is concerned with Elo in chess,
its original purpose.

The gist of this system is to track a score for each player. The
scores are defined such that the expected outcomes of games can
be computed from the scores (for example, a player with a rating
of 2400 should win 9/10 of her games against a player with a
rating of 2000). If the true outcome (of e.g.~a tournament) doesn't
match the expected outcome, then both player's scores are adjusted
towards values that would have produced the expected result. Over
time, scores thus become a more accurate reflection of player's skill,
while also allowing for players to get stronger or weaker. This
system is carefully described elsewhere, so we can just leave it
at that.

The players need not be human, and in fact this can facilitate running
many games and thereby getting accurate ratings.

The problem this paper is concerned with is that basically all chess
tournaments (whether with humans or computers or both) are between
players who know how to play chess, are interested in winning their
games, and have some reasonable level of skill. This makes it hard to
give a rating to weak players: They just lose every single game and so
tend towards a rating of $-\infty$\footnote{Some organizations don't
  even let ratings fall beneath a certain level, for example, the
  lowest possible USCF rating is 100.} Even if other comparatively
weak players existed to participate in the tournament and occasionally
lose to the player under study, it may still be difficult to
understand how this cohort performs in any absolute sense. (In
contrast we have ``the highest ever human rating was 2882,'' and
``there are 5,323 players with ratings in 2200--2299'' and ``Players
whose ratings drop below 1000 are listed on the next list as
'delisted'.''\cite{fideratings}.) It may also be the case that all
weak players always lose to all strong players, making it unclear just
how wide the performance gap between the two sets is. The goal of
this paper is to expand the dynamic range of chess ratings to span
all the way from extremely weak players to extremely strong ones, while
providing canonical reference points along the way.

\section{Elo world}

Elo World is a tournament with dozens of computer players. The
computer players include some traditional chess engines, but also many
algorithms chosen for their simplicity, as well as some designed to
be competitively bad.

The fun part is the various algorithms, so let's get into that. First,
some ground rules and guidelines:
\begin{itemize}
\item No resigning (forfeiting) or claiming a draw. The player will
  only be asked to make a move when there exists one, and it must
  choose a move in finite time. (In practice, most of the players are
  extremely fast, with the slowest ones using about one second of
  CPU time per move.)
\item The player can retain state for the game, and executes moves
  sequentially (for either the white or black pieces), but cannot have
  state meaningfully span games. For example, it is not permitted to
  do man-in-the-middle attacks~\cite{blind} or learn opponent's moves
  from previous rounds, or to get better at chess. The majority of
  players are actually completely stateless, just a partial function
  of type ${\tt position} \rightarrow {\tt move}$.
\item black and white behave the same
\item 
\end{itemize}

Ideally, such a system would be robust against adding new players,
but this is probably not the case.

% Names:
% "Stalemate"
% "Weaksauce"
% "back rank"
% "blunder"
% something with ..elo..   "elo world"
% inFIDElity
% blunderbuss
%

Interpolation, like Scoville scale.

% Dramatic failure of the "suicide king" strategy against
% CCCP.
1. d4 g5 2. Bxg5 Nc6 3. Bxe7 Kxe7 4. d5 Kd6
5. dxc6+ Kc5 6. Qd6+ Kc4 7. e4++
